---
title: "Project-6101"
author: "Sunny"
date: "2024-11-29"
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
---
```{r init, include=FALSE}
#library(ezids)
library(dplyr)
library(ggplot2)
library(corrplot)
library(car)
library(ranger)
library(pdp)
```




```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, results = "markup")
options(scipen=999, digits = 5) 
```

## 1.Data ETL


```{r}
data_raw <- read.csv("C:/Users/qluo/Box/DATS 6101/Group-Project2/Combined_SDOH_2019_2020.csv")
```

```{r}
col_indices <- c('COUNTYFIPS','YEAR','ACS_PCT_FEMALE','ACS_PCT_AGE_15_17',
                 'ACS_PCT_AGE_18_29','ACS_PCT_AGE_30_44',
                 'ACS_PCT_AIAN_NONHISP','ACS_PCT_ASIAN_NONHISP',
                 'ACS_PCT_BLACK_NONHISP',
                 'ACS_PCT_HISPANIC','ACS_PCT_NHPI_NONHISP','ACS_PCT_OTHER_NONHISP',
                 'ACS_PCT_NOT_LABOR','ACS_MEDIAN_HH_INC','ACS_PCT_HEALTH_INC_BELOW137',
                 'AHRF_UNEMPLOYED_RATE','AHRF_NURSE_MIDWIVES_RATE','CDCW_MATERNAL_DTH_RATE',
                 'POS_MEDIAN_DIST_OBSTETRICS','POS_FQHC_RATE',
                 'POS_HOSP_OBSTETRIC_RATE','POS_HOSP_MEDSURG_ICU_RATE','POS_HOSP_ED_RATE','POS_PCT_HOSP_GOV')
 
subset_data <- data_raw[,col_indices]
colnames(subset_data) <- c('countyfips','YEAR','Percent_femal_pop',"Percent_pop_15_17","Percent_pop_18_29","Percent_pop_30_44",
                           'Percent_pop_American_Indian_and_Alaska_Native','Percent_pop_Asian',
                           'Percent_pop_Black_and_African_American','Percent_pop_Hispanic',
                           'Percent_pop_Native_Hawaiian_and_Pacific_Islander','Percent_pop_other_race',
                           "Percent_pop_not_in_labor_force_16",'Median_household_income',
                           'Percent_pop_under_1.37_of_poverty_threshold',
                           "Unemployment_rate_per_100_population_16",
                           'Number_of_midwives','Maternal_mortality_rate',
                           'Median_distance_to_nearest_obstetrics_department','Numer_of_health_centers',
                           'Number_of_hospitals_with_obstetric','Number_of_hospitals_with_ICU',
                           'Number_of_hospitals_with_emergency_department','Percent_government_hospitals')

```
We can divide the variables into different categories, making them clear when showing the independent variables.

## 2. Outcome Imputation

```{r}
print(paste0("Maternal_mortality_rate is missing in ", round(sum(is.na(subset_data$Maternal_mortality_rate))/nrow(subset_data)*100,2),"% of observations."))

print(paste0("Maternal_mortality_rate is 0 in ", round(sum(subset_data$Maternal_mortality_rate==0, na.rm=T)/nrow(subset_data)*100,2),"% of observations."))

```
The highly missing nature of outcome variable *Maternal_mortality_rate* is unexpected. Also, there was no 0 maternal mortality reported in any county. We believe that ARHQ SDOH database is constructed using CDC Wonder's default output settings which suppresses deaths between 1-9 and does not output true zeros. 

We used the following procedure to impute the maternal mortality counts and collaterally the maternal mortality rates.
1. Downloaded the multiple cause maternal mortality for year 2019 and 2020 from CDC Wonder for both state and county level. We made sure that the numbers for non-missing counties matches the record in SDOH database.
2. Downloaded the all cause mortality and underlying cause that is not maternal causes at county level from CDC Wonder. By subtracting the two, we back out the underlying cause mortality due to maternal causes.
3. We used the underlying cause of mortality to distribute the gap between reported county level total mortality and state level total mortality, making sure that states with suppressed mortality has a mortality counts from the larger between underlying cause mortality and 1 to 9 and the counties with true zero maternal mortality stays true zero. 
4. We replace the imputed maternal mortality rate in the original dataframe.

### 2.1 CDC Wonder Data Processing and Imputing 

```{r}
library(readxl)
library(dplyr)
library(stringr)

# Function to clean County and State columns
clean_column <- function(column) {
  column %>%
    str_replace_all("\\*|‡|†| ", "") %>%
    str_trim()
}

# Function to handle iterative adjustments for Stage 3
adjust_imputation <- function(data) {
  # Initialize loop variables
  gap_mean <- mean(data$gap3, na.rm = TRUE)
  counter <- 0
  
  while (gap_mean != 0 & counter < 20) {
    data <- data %>%
      mutate(
        # Step 1: Identify eligible rows for imputation
        impute_elig = ifelse(
          !is.na(all_deaths_under) & is.na(maternal_multi) & between(dist3, 1, 9),
          1,
          0
        )
      ) %>%
      group_by(state_fips) %>%
      mutate(
        # Step 2: Calculate dist3_factor
        dist3_factor = ifelse(
          impute_elig == 1,
          sum(ifelse(impute_elig == 1, impute_weight, NA), na.rm = TRUE),
          NA
        )
      ) %>%
      ungroup() %>%
      mutate(
        # Step 3: Update dist3
        dist3 = ifelse(
          impute_elig == 1,
          ifelse(
            dist3 + (gap3 / dist3_factor * impute_weight) <= 9,
            dist3 + (gap3 / dist3_factor * impute_weight),
            9
          ),
          dist3
        )
      ) %>%
      group_by(state_fips) %>%
      mutate(
        # Step 4: Calculate reported_total3
        reported_total3 = sum(dist3, na.rm = TRUE)
      ) %>%
      ungroup() %>%
      mutate(
        # Step 5: Update gap3
        gap3 = maternal_multi_st - reported_total3
      )
    
    # Update loop variables
    gap_mean <- mean(data$gap3, na.rm = TRUE)
    counter <- counter + 1
  }
  
  return(data)
}

# Process Impute_2019
impute_2019 <- read_excel("MaternalMortality-addtl-data.xlsx", sheet = "Impute_2019") %>%
  mutate(
    population = as.numeric(population),
    County = clean_column(County)
  ) %>%
  filter(!is.na(County)) %>%
  mutate(
    cnty_fips = str_sub(str_sub(County, -7), 2, 6),
    state_fips = str_sub(cnty_fips, 1, 2)
  )

# Process State_2019
state_2019 <- read_excel("MaternalMortality-addtl-data.xlsx", sheet = "State_2019") %>%
  rename(maternal_multi_st = maternal_multi) %>%
  mutate(
    state = clean_column(state),
    state_fips = str_sub(str_sub(state, -3), 1, 2)
  ) %>%
  filter(!is.na(state))

# Merge and process 2019 data
merged_2019 <- left_join(impute_2019, state_2019, by = "state_fips") %>%
  group_by(state_fips) %>%
  mutate(
    # Stage 1: Calculate reported_total1 and gap1
    reported_total1 = sum(maternal_multi, na.rm = TRUE),
    gap1 = maternal_multi_st - reported_total1,
    # Stage 2: Calculate dist2, reported_total2, and gap2
    dist2 = ifelse(
      !is.na(all_deaths_under) & is.na(maternal_multi),
      pmax(maternal_under, 1),
      maternal_multi
    ),
    reported_total2 = sum(dist2, na.rm = TRUE),
    gap2 = maternal_multi_st - reported_total2
  ) %>%
  ungroup() %>%
  mutate(
    # Stage 3: Initialize dist3, gap3, and impute_weight
    dist3 = dist2,
    gap3 = gap2,
    impute_weight = maternal_under + 1
  )

# Apply Stage 3 logic to adjust dist3
merged_2019 <- adjust_imputation(merged_2019) %>%
  mutate(
    maternal_mortality_rate = dist3 / population * 100000,
    year = 2019
  ) %>%
  select(County, cnty_fips, maternal_mortality_rate, year)

# Process Impute_2020
impute_2020 <- read_excel("MaternalMortality-addtl-data.xlsx", sheet = "Impute_2020") %>%
  mutate(
    population = as.numeric(population),
    County = clean_column(County)
  ) %>%
  filter(!is.na(County)) %>%
  mutate(
    cnty_fips = str_sub(str_sub(County, -7), 2, 6),
    state_fips = str_sub(cnty_fips, 1, 2)
  )

# Process State_2020
state_2020 <- read_excel("MaternalMortality-addtl-data.xlsx", sheet = "State_2020") %>%
  rename(maternal_multi_st = maternal_multi) %>%
  mutate(
    state = clean_column(state),
    state_fips = str_sub(str_sub(state, -3), 1, 2)
  ) %>%
  filter(!is.na(state))

# Merge and process 2020 data
merged_2020 <- left_join(impute_2020, state_2020, by = "state_fips") %>%
  group_by(state_fips) %>%
  mutate(
    # Stage 1: Calculate reported_total1 and gap1
    reported_total1 = sum(maternal_multi, na.rm = TRUE),
    gap1 = maternal_multi_st - reported_total1,
    # Stage 2: Calculate dist2, reported_total2, and gap2
    dist2 = ifelse(
      !is.na(all_deaths_under) & is.na(maternal_multi),
      pmax(maternal_under, 1),
      maternal_multi
    ),
    reported_total2 = sum(dist2, na.rm = TRUE),
    gap2 = maternal_multi_st - reported_total2
  ) %>%
  ungroup() %>%
  mutate(
    # Stage 3: Initialize dist3, gap3, and impute_weight
    dist3 = dist2,
    gap3 = gap2,
    impute_weight = maternal_under + 1
  )

# Apply Stage 3 logic to adjust dist3
merged_2020 <- adjust_imputation(merged_2020) %>%
  mutate(
    maternal_mortality_rate = dist3 / population * 100000,
    year = 2020
  ) %>%
  select(County, cnty_fips, maternal_mortality_rate, year)

# Combine 2019 and 2020 data
combined <- bind_rows(merged_2019, merged_2020)

# View final combined data
print(head(combined))
```

### 2.2 Merge Imputed Outcome to SDOH Data

```{r}
# Ensure COUNTYFIPS in subset_data is formatted as a 5-digit string
subset_data <- subset_data %>%
  mutate(
    countyfips = sprintf("%05d", as.integer(countyfips))
  )

# Join subset_data with combined on YEAR and COUNTYFIPS/cnty_fips
merged_data <- subset_data %>%
  left_join(
    combined,
    by = c("YEAR" = "year", "countyfips" = "cnty_fips")
  )

# Fill missing Maternal_mortality_rates in subset_data with values from combined
subset_data <- merged_data %>%
  mutate(
    Maternal_mortality_rate = ifelse(
      is.na(Maternal_mortality_rate),
      maternal_mortality_rate,
      Maternal_mortality_rate
    )
  ) %>%
  select(-maternal_mortality_rate,-countyfips, -YEAR,-County) # Remove unnecessary columns if desired
# View the updated subset_data
print(head(subset_data))
```
### 2.3 KNN Imputation of Other Variables
```{r}
if (!require("VIM")) install.packages("VIM")
library(VIM)



outcome_column <- subset_data$Maternal_mortality_rate
subset_data_no_outcome <- subset_data %>%
  select(-Maternal_mortality_rate)

# Perform KNN imputation on the predictors
subset_data_imputed <- kNN(subset_data_no_outcome, k = 5, imp_var = FALSE)

# Add the outcome column back
subset_data_imputed$outcome <- outcome_column

# Apply KNN imputation
subset_data_imputed <- kNN(subset_data, k = 5, imp_var = FALSE)



# View the first few rows of the imputed dataframe
head(subset_data_imputed)

subset_data=subset_data_imputed
```
## 3. Descriptive Analysis of Maternal Mortality Rates

It shows non-normal distribution.

```{r}
hist(subset_data$Maternal_mortality_rate,
     xlim = c(0, 5),  # Adjust the x-axis to a smaller range
     breaks = 200,      # Increase the number of bins for better granularity
     col = "lightblue", # Add color to make it visually appealing
     main = "Histogram of Maternal Mortality Rate",
     xlab = "Maternal Mortality Rate",
     ylab = "Frequency"
     )

```

The bulk is 0.There are `r sum(subset_data$Maternal_mortality_rate == 0)` of 0.

## 4. Linear Regression

We ran linear regression in this section.

### 4.1 VIF for multicollinearity
```{r}
model <- lm(Maternal_mortality_rate ~ ., data=subset_data)
vif_values <- vif(model)
print(vif_values)

```
The VIF values are relatively high for `Median_household_income` and `Percent_pop_under_1.37_of_poverty_threshold`. Doing the correlation test to check the multicolinearity.    

```{r}

# Spearman correlation
cor.test(subset_data$Median_household_income, subset_data$Percent_pop_under_1.37_of_poverty_threshold, method = "spearman")

```
The coefficients for the two variables are higher than 0.8. Drop `Percent_pop_under_1.37_of_poverty_threshold` and run the final model.

### 4.2 Linear Regression Model
```{r}
# Fit a model excluding Percent_pop_under_1.37_of_poverty_threshold
model_1 <- lm(Maternal_mortality_rate ~ . - Percent_pop_under_1.37_of_poverty_threshold, data = subset_data)

# View summary
summary(model_1)

```
The coefficients for `Percent_pop_American_Indian_and_Alaska_Native` and `Percent_pop_Black_and_African_American` are statistically significant and positive, indicating that counties with higher percentages of American Indian and Black populations have higher maternal mortality rates. Highlights racial disparities in maternal health outcomes, driven by systemic inequities, implicit bias in healthcare, and differences in socioeconomic conditions. Structural factors like unequal access to healthcare, chronic stress from discrimination, and higher rates of comorbidities may contribute to this disparity.

The coefficient for `Median_household_income` is statistically significant and negative, suggesting that higher median household incomes are associated with lower maternal mortality rates. Wealthier counties likely have better healthcare infrastructure, more resources for maternal care, and higher access to prenatal and postpartum services. Higher income levels often correlate with better health literacy, healthier lifestyles, and access to private healthcare services.

The coefficient for `Number_of_hospitals_with_ICU` is statistically significant and negative, showing that Counties with more hospitals equipped with ICU facilities tend to have lower maternal mortality rates.ICU availability indicates the capacity to handle severe maternal complications (e.g., hemorrhage, sepsis) effectively.Hospitals with ICUs may also have higher overall quality of care, better-trained staff, and access to advanced medical technology.

The coefficient for `Percent_femal_pop` is negative and statistically significant, indicating a higher percentage of females in the population is associated with a lower maternal mortality rate. This may indicate that communities with a higher proportion of females have better infrastructure and resources for women's health, or more gender-focused policies that improve maternal health outcomes.
It could also reflect cultural or social dynamics where higher female representation leads to stronger support systems for maternal care.

The coefficient for `Percent_pop_30_44` is statistically significant and negative, indicating that counties with a higher percentage of women aged 30-44 have lower maternal mortality rates. This group may represent the most biologically optimal age range for pregnancy compared to older groups (e.g., >44 years), leading to lower risks of complications.Women in this age group may also have better access to prenatal care or more stable socioeconomic conditions compared to younger women (e.g., <30 years).This may reflect the fact that, compared to the base group (women older than 44), this age group experiences fewer pregnancy-related complications.

The coefficient for `Number_of_hospitals_with_obstetric` is statistically significant and positive. Areas with more hospitals offering obstetric services often attract higher-risk pregnant women who require specialized care. These women may already face greater complications, leading to higher maternal mortality rates in these hospitals.This aligns well with healthcare patterns, where high-risk cases tend to cluster around centers of excellence or specialized facilities.


### 4.3 Diagnostic Plots

```{r}
par(mfrow = c(2, 2))  # Set up a 2x2 plot grid
plot(model_1)
```

```{r}
residuals <- residuals(model_1)
hist(residuals, 
     xlim=c(-5,5),
     breaks=100,
     main = "Histogram of Residuals", xlab = "Residuals")
```
The distribution of residual is slightly skewed. The skewness is minor and the sample size is relatively large. The central limit theorem can mitigate its impact on inference.

## 5. Random Forest Regression
We ran the random forest regressions in this section.
### 5.1 Hyperparameter Tuning using MSE as Metrics

```{r}
# Hyperparameter tuning for random forest regression tree 
# Define the tuning grid: include additional parameters
tuning_grid <- expand.grid(
  num.trees = c(100, 200, 300, 500),
  mtry = 1:(ncol(subset_data) - 1),
  splitrule = c("variance",'extratrees','maxstat'),    # Split rule for regression
  min.node.size = c(1, 5, 10),
  max.depth = c(0, 5, 10, 20)
)

# Randomly sample 60 combinations from the grid
set.seed(456)
random_combinations <- tuning_grid[sample(nrow(tuning_grid), 60), ]

# Cross-validation function
cross_validate <- function(data, formula, params, k = 5) {
  folds <- sample(1:k, nrow(data), replace = TRUE)
  scores <- numeric(k)
  
  for (i in 1:k) {
    # Split data into training and validation sets
    train_data <- data[folds != i, ]
    val_data <- data[folds == i, ]
    
    # Fit the model
    model <- ranger(
      formula = formula,
      data = train_data,
      splitrule = params$splitrule,
      num.trees = params$num.trees,
      mtry = params$mtry,
      min.node.size = params$min.node.size,
      max.depth = params$max.depth,
      num.threads=16
    )
    
    # Predict and calculate RMSE
    predictions <- predict(model, val_data)$predictions
    scores[i] <- sqrt(mean((val_data$Maternal_mortality_rate - predictions)^2))
  }
  
  # Return mean RMSE across folds
  mean(scores)
}

# Perform random search
results <- data.frame(random_combinations, RMSE = NA)

for (i in 1:nrow(random_combinations)) {
  if (i%%50==0)(
    cat(paste(".    ",i,"\r"))
  )
  else{
    cat(".")
  }
  params <- random_combinations[i, ]
  results$RMSE[i] <- cross_validate(
    data = subset_data,
    formula = Maternal_mortality_rate ~ .,
    params = params
  )
}
```
Best Model from Hyperparameter Tuning

```{r}

# Find the best combination
best_model <- results[which.min(results$RMSE), ]
best_model

```

### 5.2 Refit the final Random Forest model

```{r}
# Fit the Final Tree-Based Model
library(randomForest)
set.seed(123)
final_rf_model <- ranger(
  Maternal_mortality_rate ~ .,
  data = subset_data,
  num.trees = best_model$num.trees,
  splitrule = best_model$splitrule,
  mtry = best_model$mtry,
  min.node.size = best_model$min.node.size,
  max.depth = best_model$max.depth,
  num.threads=16,
  importance='permutation'
  )


```

```{r}
print(final_rf_model)
```

### 5.3 Parameter Importance 
```{r}
# Extract the importance metrics as a named vector
importance_metrics <- final_rf_model$variable.importance

# Convert to a data frame for easier manipulation
importance_df <- data.frame(
  Variable = names(importance_metrics),
  `%IncMSE` = importance_metrics
)

# Sort by `%IncMSE` in descending order
sorted_by_mse <- importance_df[order(-importance_df$`X.IncMSE`), ]

# View the sorted importance
print(sorted_by_mse)
```

The column `%IncMSE` stands for percentage increase in Mean Squared Error.
Measures how much the model’s prediction error increases when a particular variable is randomly permuted while keeping all other variables unchanged..
Higher values indicate that the variable is important for the model's predictive accuracy.It reflects the variable's contribution to the overall model performance.

```{r}
# Adjust margins for longer labels
par(mar = c(10, 5, 2, 2))  # Increase bottom margin

# Bar plot with adjusted margins
barplot(sorted_by_mse[, "X.IncMSE"],
        names.arg = rownames(sorted_by_mse),
        las = 2,
        col = "skyblue",
        main = "Variable Importance by %IncMSE",
        ylab = "%IncMSE",
        cex.names = 0.5)
```


### 5.4 Partial Dependence of Key Features

```{r}
# Partial dependence for multiple features
features <- rownames(sorted_by_mse)

# Loop to generate and print plots

for (feature in features) {
  pd <- partial(
    object = final_rf_model,
    pred.var = feature,
    train = subset_data,
    pred.fun = function(object, newdata) {
      predict(object, data = newdata)$predictions
    }
  )
  
  # Convert partial dependence to a data frame
  pd_df <- as.data.frame(pd)
  # Apply spline smoothing
  spline_fit <- smooth.spline(pd_df[,feature], pd_df$yhat, spar = 0.7)  # spar = smoothing parameter
  
  # Convert spline results to a data frame for ggplot
  spline_df <- data.frame(
    feature = spline_fit$x,
    yhat_smooth = spline_fit$y
  )
  colnames(spline_df)[1] <- feature
  # Create a proper ggplot
  print(ggplot(spline_df, aes_string(x = feature, y = "yhat_smooth")) +
    geom_line() +
    ggtitle(paste("Partial Dependence of", feature)) +
    xlab(feature) +
    ylab("Predicted Maternal Mortality Rate") +
    theme_minimal())
}
```
Partial Dependence of Percent_pop_18_29:

The maternal mortality rate increases steeply as the percentage of the population aged 18–29 rises to approximately 20%.
This suggests that a higher percentage of younger populations may contribute to maternal mortality initially, likely due to increased birth rates in this group. 

Partial Dependence of Percent_pop_Black_and_African_American and Percent_pop_American_Indian_and_Alaska_Native:

The maternal mortality rate steadily increases as the percentage of these group population rises.
This aligns with systemic inequities and healthcare access disparities often faced by this demographic, emphasizing the need for targeted interventions.

Partial Dependence of Percent_pop_under_1.37_of_poverty_threshold:

Maternal mortality rises as the percentage of people living below 1.37 times the poverty threshold increases, but the growth slows after around 40%.
This indicates that economic hardship significantly impacts maternal mortality, although the incremental effect diminishes at higher poverty levels.

Partial Dependence of Median_household_income:

The maternal mortality rate declines steeply as median household income increases.
This reinforces the association between higher socioeconomic status and improved maternal health outcomes, highlighting the importance of reducing economic disparities.

Partial Dependence of Median_distance_to_nearest_obstetrics_department:

Maternal mortality initially increases as the distance to the nearest obstetrics department grows but plateaus after 100 km.
This indicates that limited access to obstetric care contributes to higher maternal mortality, but in rural areas with very high distances, the overall population and number of births may be lower, potentially mitigating the impact.

1. For Percent_pop_18_29 (Young Population):
Finding: Maternal mortality increases as the percentage of the population aged 18–29 rises but plateaus after a certain threshold.
Solutions:
Enhance prenatal and perinatal care: Ensure accessible and affordable healthcare for younger mothers to address potential complications during pregnancy and childbirth.
Promote family planning education: Equip young women with knowledge about reproductive health and safe motherhood practices.
Expand community health programs: Develop support groups and programs tailored for younger mothers to address unique risks they face.

2. For Percent_pop_Black_and_African_American (Racial Disparities):
Finding: Maternal mortality increases with a higher percentage of the Black or African American population.
Solutions:
Address healthcare access disparities: Establish or enhance healthcare facilities in underserved areas with a high percentage of minority populations.
Implement culturally tailored healthcare programs: Train healthcare providers in cultural competency and implicit bias reduction to provide equitable care.
Community outreach and engagement: Build trust within these communities by involving local leaders and organizations in healthcare initiatives.

3. For Percent_pop_under_1.37_of_poverty_threshold (Economic Hardship):
Finding: Maternal mortality rises significantly with higher poverty levels but stabilizes beyond a certain point.
Solutions:
Provide financial assistance programs: Expand Medicaid coverage, maternal health subsidies, and food assistance programs for low-income families.
Improve education and job opportunities: Invest in education and workforce development programs to lift families out of poverty in the long term.
Increase access to affordable healthcare: Reduce or eliminate out-of-pocket expenses for maternal health services, especially for low-income populations.

4. For Median_household_income (Income Inequality):
Finding: Higher median household income is associated with lower maternal mortality.
Solutions:
Strengthen social safety nets: Ensure income support programs and tax relief policies target vulnerable populations.
Expand healthcare coverage: Provide universal maternal healthcare to ensure that income disparities do not impact maternal health outcomes.
Target rural economic development: Create jobs and incentivize businesses to operate in underserved areas to raise household income levels.

5. For Median_distance_to_nearest_obstetrics_department (Access to Care):
Finding: Maternal mortality increases with greater distances to obstetrics departments, particularly up to 100 km.
Solutions:
Build more healthcare facilities: Increase the number of hospitals and clinics with obstetrics services in rural and remote areas.
Telemedicine and mobile clinics: Use telemedicine platforms and mobile health units to provide care to women in areas with limited access.
Subsidize transportation costs: Implement programs that cover transportation to medical appointments for pregnant women in remote locations.



1. Socioeconomic Factors
Findings:

Maternal mortality is positively associated with a higher percentage of the population living under 1.37 times the poverty threshold.
Higher median household income is associated with lower maternal mortality.
Solutions:

Economic Support:
Provide financial assistance programs, such as Medicaid expansion, subsidies for prenatal care, and food assistance for low-income families.
Strengthen social safety nets like tax relief policies and cash transfer programs.
Job and Education Opportunities:
Invest in education and workforce development programs to help families move out of poverty.
Promote rural economic development initiatives to increase household income levels.
Affordable Healthcare:
Reduce out-of-pocket healthcare expenses and ensure affordable maternal health services for low-income families.
2. Demographic Trends
Findings:

Higher maternal mortality is associated with a higher percentage of young adults aged 18–29 in the population.
Black or African American and American Indian populations have higher maternal mortality rates, indicating systemic inequities.
Solutions:

Young Population:
Enhance prenatal and perinatal care for younger mothers through accessible healthcare services.
Promote family planning education to empower younger women with knowledge about reproductive health and safe motherhood practices.
Develop community-based support programs tailored to the unique risks faced by younger mothers.
Racial Disparities:
Address healthcare access gaps in minority-dense areas by establishing or enhancing healthcare facilities.
Implement culturally tailored maternal health programs and train providers in cultural competency to ensure equitable care.
Engage with community leaders and organizations to build trust and raise awareness about maternal health services.
3. Healthcare Access
Findings:

Maternal mortality increases with greater distances to the nearest obstetrics department (up to 100 km).
More hospitals with ICUs are associated with a lower maternal mortality rate.
More hospitals with obstetric departments are positively correlated with maternal mortality, potentially reflecting referral bias.
Solutions:

Improve Access to Care:
Build more healthcare facilities with obstetric and ICU services in rural and underserved areas.
Subsidize transportation costs for pregnant women in remote locations to access obstetric care.
Leverage Telemedicine and Mobile Clinics:
Use telemedicine platforms and deploy mobile health units to reach women in remote areas.
Optimize Referral Systems:
Analyze referral patterns to ensure high-risk patients are appropriately routed to hospitals with obstetric and ICU services.
Strengthen coordination between hospitals to balance the patient load and provide timely care.
