---
title: "GROUP PROJECT 2 REPORT "
author: "TEAM7 - AKSHIT , SUNNY"
date: "2024-11-29"
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
---

#INTRODUCTION 

This Project focuses on understanding the socioeconomic and healthcare-related factors behind maternal death rates.BY analyzing a range of data, we aimed to uncover patters and identify areas for improvement to address this critical challenge


#DATASET OVERVIEW

##What do we know about this dataset?

The Social Determinants of Health dataset was developed by the Agency for Healthcare Research
and Quality (AHRQ). It consists of county-level data across the United States, covering various
socio-economic and healthcare factors. It is used significantly in public health research to determine how social determinants impact health outcomes.

## How was the information gathered?

Data was collected from multiple sources, including national surveys, administrative records, and government databases, ensuring a wide representation of social determinants. 

 
## What are the limitations of the dataset

1.Missing Data
2.Since the data is aggregated at the county level, it may not reflect
individual or household-level variations

#Research Question

What socioeconomic and healthcare-related factors significantly influence maternal mortality rates?.

Our Research question emphasizes inference.

#Model Selection 

1. Linear Regression : Selected for inference to determine the relationship between variables.

2. Random Forest Regression : Used for prediction to evaluate feature importance and capture nonlinear        relationships.

#DEPENDENT VARIABLE

Maternal Mortality Rate

#INDEPENDENT VARIABLE

DEMOGRAPHIC VARIABLES :-

Percentage of Female Population
Percentage of Population Aged 15–17, 18–29, and 30–44
Percentage of American Indian/Alaska Native Population
Percentage of Asian Population
Percentage of Black/African American Population
Percentage of Hispanic Population
Percentage of Native Hawaiian/Pacific Islander Population
Percentage of Other Races

SOCIOECONOMIC VARIABLES:-

Median Household Income
Percentage Below the Poverty Threshold (1.37x)
Unemployment Rate (16+)
Percentage Not in the Labor Force (16+)

HEALTHCARE ACCESS VARIABLES:-

Median Household Income
Percentage Below the Poverty Threshold (1.37x)
Unemployment Rate (16+)
Percentage Not in the Labor Force (16+)

#DATAPREPROCESSING

##DATA EXTRACTION AND CLEANING 

The dataset was read from a csv file and relevant columns were extracted for analysis.




```{r init, include=FALSE}
#library(ezids)

library(dplyr)
library(ggplot2)
library(corrplot)
library(car)
library(ranger)
library(pdp)
```




```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, results = "markup")
options(scipen=999, digits = 5) 
```


```{r}
data_raw <- read.csv("C:\\Users\\91970\\OneDrive\\Desktop\\GP 2 DS\\Combined_SDOH_2019_2020.csv")
```

```{r}
col_indices <- c('COUNTYFIPS','YEAR','ACS_PCT_FEMALE','ACS_PCT_AGE_15_17',
                 'ACS_PCT_AGE_18_29','ACS_PCT_AGE_30_44',
                 'ACS_PCT_AIAN_NONHISP','ACS_PCT_ASIAN_NONHISP',
                 'ACS_PCT_BLACK_NONHISP',
                 'ACS_PCT_HISPANIC','ACS_PCT_NHPI_NONHISP','ACS_PCT_OTHER_NONHISP',
                 'ACS_PCT_NOT_LABOR','ACS_MEDIAN_HH_INC','ACS_PCT_HEALTH_INC_BELOW137',
                 'AHRF_UNEMPLOYED_RATE','AHRF_NURSE_MIDWIVES_RATE','CDCW_MATERNAL_DTH_RATE',
                 'POS_MEDIAN_DIST_OBSTETRICS','POS_FQHC_RATE',
                 'POS_HOSP_OBSTETRIC_RATE','POS_HOSP_MEDSURG_ICU_RATE','POS_HOSP_ED_RATE','POS_PCT_HOSP_GOV')
 
subset_data <- data_raw[,col_indices]
colnames(subset_data) <- c('countyfips','YEAR','Percent_femal_pop',"Percent_pop_15_17","Percent_pop_18_29","Percent_pop_30_44",
                           'Percent_pop_American_Indian_and_Alaska_Native','Percent_pop_Asian',
                           'Percent_pop_Black_and_African_American','Percent_pop_Hispanic',
                           'Percent_pop_Native_Hawaiian_and_Pacific_Islander','Percent_pop_other_race',
                           "Percent_pop_not_in_labor_force_16",'Median_household_income',
                           'Percent_pop_under_1.37_of_poverty_threshold',
                           "Unemployment_rate_per_100_population_16",
                           'Number_of_midwives','Maternal_mortality_rate',
                           'Median_distance_to_nearest_obstetrics_department','Numer_of_health_centers',
                           'Number_of_hospitals_with_obstetric','Number_of_hospitals_with_ICU',
                           'Number_of_hospitals_with_emergency_department','Percent_government_hospitals')

```


##Outcome Imputation

The Maternal Mortality Rate was highly missing. Imputation was performed using a multi-step method involving external CDC Wonder data. 

Missing rates were imputed using underlying cause mortality and state-level statistics to estimate suppressed values for counties.

```{r}
print(paste0("Maternal_mortality_rate is missing in ", round(sum(is.na(subset_data$Maternal_mortality_rate))/nrow(subset_data)*100,2),"% of observations."))

print(paste0("Maternal_mortality_rate is 0 in ", round(sum(subset_data$Maternal_mortality_rate==0, na.rm=T)/nrow(subset_data)*100,2),"% of observations."))

```


```{r}
library(readxl)
library(dplyr)
library(stringr)


clean_column <- function(column) {
  column %>%
    str_replace_all("\\*|‡|†| ", "") %>%
    str_trim()
}

adjust_imputation <- function(data) {
  
  gap_mean <- mean(data$gap3, na.rm = TRUE)
  counter <- 0
  
  while (gap_mean != 0 & counter < 20) {
    data <- data %>%
      mutate(
        
        impute_elig = ifelse(
          !is.na(all_deaths_under) & is.na(maternal_multi) & between(dist3, 1, 9),
          1,
          0
        )
      ) %>%
      group_by(state_fips) %>%
      mutate(
        
        dist3_factor = ifelse(
          impute_elig == 1,
          sum(ifelse(impute_elig == 1, impute_weight, NA), na.rm = TRUE),
          NA
        )
      ) %>%
      ungroup() %>%
      mutate(
       
        dist3 = ifelse(
          impute_elig == 1,
          ifelse(
            dist3 + (gap3 / dist3_factor * impute_weight) <= 9,
            dist3 + (gap3 / dist3_factor * impute_weight),
            9
          ),
          dist3
        )
      ) %>%
      group_by(state_fips) %>%
      mutate(
        
        reported_total3 = sum(dist3, na.rm = TRUE)
      ) %>%
      ungroup() %>%
      mutate(
        
        gap3 = maternal_multi_st - reported_total3
      )
    
    
    gap_mean <- mean(data$gap3, na.rm = TRUE)
    counter <- counter + 1
  }
  
  return(data)
}


impute_2019 <- read_excel("C:\\Users\\91970\\OneDrive\\Desktop\\GP 2 DS\\MaternalMortality-addtl-data.xlsx", sheet = "Impute_2019") %>%
  
  mutate(
    population = as.numeric(population),
    County = clean_column(County)
  ) %>%
  filter(!is.na(County)) %>%
  mutate(
    cnty_fips = str_sub(str_sub(County, -7), 2, 6),
    state_fips = str_sub(cnty_fips, 1, 2)
  )


state_2019 <- read_excel("C:\\Users\\91970\\OneDrive\\Desktop\\GP 2 DS\\MaternalMortality-addtl-data.xlsx", sheet = "State_2019") %>%
  rename(maternal_multi_st = maternal_multi) %>%
  mutate(
    state = clean_column(state),
    state_fips = str_sub(str_sub(state, -3), 1, 2)
  ) %>%
  filter(!is.na(state))


merged_2019 <- left_join(impute_2019, state_2019, by = "state_fips") %>%
  group_by(state_fips) %>%
  mutate(
    
    reported_total1 = sum(maternal_multi, na.rm = TRUE),
    gap1 = maternal_multi_st - reported_total1,
    
    dist2 = ifelse(
      !is.na(all_deaths_under) & is.na(maternal_multi),
      pmax(maternal_under, 1),
      maternal_multi
    ),
    reported_total2 = sum(dist2, na.rm = TRUE),
    gap2 = maternal_multi_st - reported_total2
  ) %>%
  ungroup() %>%
  mutate(
   
    dist3 = dist2,
    gap3 = gap2,
    impute_weight = maternal_under + 1
  )


merged_2019 <- adjust_imputation(merged_2019) %>%
  mutate(
    maternal_mortality_rate = dist3 / population * 100000,
    year = 2019
  ) %>%
  select(County, cnty_fips, maternal_mortality_rate, year)


impute_2020 <- read_excel("C:\\Users\\91970\\OneDrive\\Desktop\\GP 2 DS\\MaternalMortality-addtl-data.xlsx", sheet = "Impute_2020") %>%
  mutate(
    population = as.numeric(population),
    County = clean_column(County)
  ) %>%
  filter(!is.na(County)) %>%
  mutate(
    cnty_fips = str_sub(str_sub(County, -7), 2, 6),
    state_fips = str_sub(cnty_fips, 1, 2)
  )


state_2020 <- read_excel("C:\\Users\\91970\\OneDrive\\Desktop\\GP 2 DS\\MaternalMortality-addtl-data.xlsx", sheet = "State_2020") %>%
  rename(maternal_multi_st = maternal_multi) %>%
  mutate(
    state = clean_column(state),
    state_fips = str_sub(str_sub(state, -3), 1, 2)
  ) %>%
  filter(!is.na(state))


merged_2020 <- left_join(impute_2020, state_2020, by = "state_fips") %>%
  group_by(state_fips) %>%
  mutate(
    
    reported_total1 = sum(maternal_multi, na.rm = TRUE),
    gap1 = maternal_multi_st - reported_total1,
    
    dist2 = ifelse(
      !is.na(all_deaths_under) & is.na(maternal_multi),
      pmax(maternal_under, 1),
      maternal_multi
    ),
    reported_total2 = sum(dist2, na.rm = TRUE),
    gap2 = maternal_multi_st - reported_total2
  ) %>%
  ungroup() %>%
  mutate(
    
    dist3 = dist2,
    gap3 = gap2,
    impute_weight = maternal_under + 1
  )


merged_2020 <- adjust_imputation(merged_2020) %>%
  mutate(
    maternal_mortality_rate = dist3 / population * 100000,
    year = 2020
  ) %>%
  select(County, cnty_fips, maternal_mortality_rate, year)


combined <- bind_rows(merged_2019, merged_2020)


print(head(combined))
```



```{r}

subset_data <- subset_data %>%
  mutate(
    countyfips = sprintf("%05d", as.integer(countyfips))
  )


merged_data <- subset_data %>%
  left_join(
    combined,
    by = c("YEAR" = "year", "countyfips" = "cnty_fips")
  )


subset_data <- merged_data %>%
  mutate(
    Maternal_mortality_rate = ifelse(
      is.na(Maternal_mortality_rate),
      maternal_mortality_rate,
      Maternal_mortality_rate
    )
  ) %>%
  select(-maternal_mortality_rate,-countyfips, -YEAR,-County)

print(head(subset_data))
```
##KNN Imputation for Predictors

The KNN method was applied to impute missing values in predictor variables. 

This ensured that missing data points did not bias the analysis while maintaining data structure

```{r}
if (!require("VIM")) install.packages("VIM")
library(VIM)



outcome_column <- subset_data$Maternal_mortality_rate
subset_data_no_outcome <- subset_data %>%
  select(-Maternal_mortality_rate)


subset_data_imputed <- kNN(subset_data_no_outcome, k = 5, imp_var = FALSE)


subset_data_imputed$outcome <- outcome_column


subset_data_imputed <- kNN(subset_data, k = 5, imp_var = FALSE)




head(subset_data_imputed)

subset_data=subset_data_imputed
```

##VIF

We calculated VIF for all factors to detect multicollinearity. 

Median Household Income and Percentage Population Under 1.37 times the Poverty Threshold exhibited the highest multicollinearity.

Most variables such as Number of Health Centres etc had VIF below 2, indication minimal collinearity. 



```{r}
model <- lm(Maternal_mortality_rate ~ ., data=subset_data)
vif_values <- vif(model)
print(vif_values)

```
 
##Spearman Correlation 

We performed the Spearman correlation test to assess the strength of the relationship between Median Household Income and Percent Population Under 1.37 times poverty threshold.

The Spearman's rank correlation was -0.8787, indicating a strong negative correlation between the two variables.

So based on the correlation results, we dropped the Percent Population Under 1.37 times the poverty threshold variables from the final regression model.

```{r}


cor.test(subset_data$Median_household_income, subset_data$Percent_pop_under_1.37_of_poverty_threshold, method = "spearman")

```

#Descriptive Analysis of Maternal Mortality Rates

The Histogram is right skewed showing that majority of the regions have low maternal deaths. 

Small number of regions exhibit higher rates, creating a long right tail.

So it shows that Maternal Deaths are rare in most places and only a few places experience high maternal deaths.

```{r}
hist(subset_data$Maternal_mortality_rate,
     xlim = c(0, 5),  
     breaks = 200,     
     col = "lightblue", 
     main = "Histogram of Maternal Mortality Rate",
     xlab = "Maternal Mortality Rate",
     ylab = "Frequency"
     )

```

#Linear Regression Model

We performed linear regression to understand how demographic, socioeconomic and healthcare factors relate to Maternal Mortality Rates.

```{r}
# Fit a model excluding Percent_pop_under_1.37_of_poverty_threshold
model_1 <- lm(Maternal_mortality_rate ~ . - Percent_pop_under_1.37_of_poverty_threshold, data = subset_data)

# View summary
summary(model_1)

```

##FINDINGS
From the results we see that:-

1. Coefficients for `Percent_pop_American_Indian_and_Alaska_Native` and `Percent_pop_Black_and_African_American` are positive and significant,showing that counties with higher percentages of American Indian and Black populations tend to have higher maternal mortality rates.

2.The negative and significant coefficient for Median household income indicates that higher household incomes are linked to lower maternal mortality rates.

3.The coefficient for Number_of_hospitals_with_ICU is negative and significant, indicating that counties with more ICU-equipped hospitals tend to have lower maternal mortality rates.

4.The negative and significant coefficient for Percent_femal_pop suggests that a higher percentage of females in the population is linked to lower maternal mortality rates.

5.The positive and significant coefficient for Number_of_hospitals_with_obstetric indicates that areas with more obstetric hospitals often serve higher-risk pregnancies, likely due to specialized care availability.

##MODEL PREFORMANCE

1. Multiple R Squared values is 0.0527, meaning the model explains 5.27% of the variation in maternal mortality rates

2. The Adjusted R-Squares is 0.0498 which is low, accounting for the number of predictors in the model.

3. The F-statistic is 17.9 showing that predictors have a significant affect on maternal mortality rates.

##Limitations

1. Low R-Squared Values 

2. Risk of overfitting







```{r}
par(mfrow = c(2, 2))  # Set up a 2x2 plot grid
plot(model_1)
```

##Model Diagnostics 

1. Residuals vs Fitted

- From the plot we can see that the residuals are scattered randomly, but some patterns suggest non linearity or uneven variance

2. Q-Q plot 

- From the plot we can see that points deviate from the line at the tails. Showing that the residuals are not normally distributed.

3. Scale-Location Plot

- From the plot we can see that residuals are scattered randomly but there are a few signs of uneven spread.

4. Residuals vs Leverage 
- From the plot we can see that a few points have high data leverage and cook's distance, meaning they could strongly affect the model.

-> These plots show that the model performs well but can be improved by addressing normality issues.


# Random Forest Regression

We performed Random Forest Regression to complement the inear regression model by addressing it's limitations and providing additional insights.

## Hyperparameter Tuning

- We created a tuning grid to explore combinations of key hyperparameters 

1.num.trees:- Number of trees in the forest: 100, 200,300,500.

2.mtry:- Number of features randomly selected at each split from 1 to total number of predictors.

3.splitrule:- Methods for splliting nodes: Variance,Extratrees,maxstat.

4.min.node.size:- Minimum number of data points required in a terminal node: 1,5,10.

5.max.depths:- Maximum depth of a tree: 0, 5, 10, 20.

- A 5-fold cross validation was used to evaluate each combination

- The dataset was split into 5 folds, with 4 folds used for training and 1 for validation, iteratively

- For each combination, average RMSE was recorded.

- The best performing combination was selected based on the lowest RMSE.

```{r}
 

tuning_grid <- expand.grid(
  num.trees = c(100, 200, 300, 500),
  mtry = 1:(ncol(subset_data) - 1),
  splitrule = c("variance",'extratrees','maxstat'),    
  min.node.size = c(1, 5, 10),
  max.depth = c(0, 5, 10, 20)
)


set.seed(456)
random_combinations <- tuning_grid[sample(nrow(tuning_grid), 60), ]


cross_validate <- function(data, formula, params, k = 5) {
  folds <- sample(1:k, nrow(data), replace = TRUE)
  scores <- numeric(k)
  
  for (i in 1:k) {
    
    train_data <- data[folds != i, ]
    val_data <- data[folds == i, ]
    
  
    model <- ranger(
      formula = formula,
      data = train_data,
      splitrule = params$splitrule,
      num.trees = params$num.trees,
      mtry = params$mtry,
      min.node.size = params$min.node.size,
      max.depth = params$max.depth,
      num.threads=16
    )
    
    
    predictions <- predict(model, val_data)$predictions
    scores[i] <- sqrt(mean((val_data$Maternal_mortality_rate - predictions)^2))
  }
  

  mean(scores)
}


results <- data.frame(random_combinations, RMSE = NA)

for (i in 1:nrow(random_combinations)) {
  if (i%%50==0)(
    cat(paste(".    ",i,"\r"))
  )
  else{
    cat(".")
  }
  params <- random_combinations[i, ]
  results$RMSE[i] <- cross_validate(
    data = subset_data,
    formula = Maternal_mortality_rate ~ .,
    params = params
  )
}

best_model <- results[which.min(results$RMSE), ]
best_model

```

- Best model for tuning : 

1. num.trees: 100
2. mtry : 21
3. splitrule : maxstat
4. min.node.size : 1
5. max.depth : 20
6. Best RMSE: 1.3184

## Refitting the Random Forest Model

- We refitted the Random Forest Model using the Best Hyperparameters obtained from the tuning process.

- This Final Random Forest Model provides robust predictions of maternal mortality rates and identifies key facotrs.

```{r}

library(randomForest)
set.seed(123)
final_rf_model <- ranger(
  Maternal_mortality_rate ~ .,
  data = subset_data,
  num.trees = best_model$num.trees,
  splitrule = best_model$splitrule,
  mtry = best_model$mtry,
  min.node.size = best_model$min.node.size,
  max.depth = best_model$max.depth,
  num.threads=16,
  importance='permutation'
  )

print(final_rf_model)

```

## Parameter Importance

- The Random Forest Model identified the most important factors influencing maternal mortality rates.

```{r}

importance_metrics <- final_rf_model$variable.importance


importance_df <- data.frame(
  Variable = names(importance_metrics),
  `%IncMSE` = importance_metrics
)


sorted_by_mse <- importance_df[order(-importance_df$`X.IncMSE`), ]


print(sorted_by_mse)


par(mar = c(10, 5, 2, 2))  # Increase bottom margin

# Bar plot with adjusted margins
barplot(sorted_by_mse[, "X.IncMSE"],
        names.arg = rownames(sorted_by_mse),
        las = 2,
        col = "skyblue",
        main = "Variable Importance by %IncMSE",
        ylab = "%IncMSE",
        cex.names = 0.5)
```

From the results we see that : 

1. Age group 18-29 is the most influencial. Showing higher maternal risk 

2. Higher income reduces maternal mortality rates.

3. Percentage of population below 1.37times the poverty threshold have high maternal mortality rates

4. Longer distance to healthcare increases the risk of maternal mortality rates

5. More heealth centres help reduce Maternal MOrtality Rates.

##Partial Dependence of Key Features

```{r}
# Partial dependence for multiple features
features <- rownames(sorted_by_mse)

# Loop to generate and print plots

for (feature in features) {
  pd <- partial(
    object = final_rf_model,
    pred.var = feature,
    train = subset_data,
    pred.fun = function(object, newdata) {
      predict(object, data = newdata)$predictions
    }
  )
  
  # Convert partial dependence to a data frame
  pd_df <- as.data.frame(pd)
  # Apply spline smoothing
  spline_fit <- smooth.spline(pd_df[,feature], pd_df$yhat, spar = 0.7)  # spar = smoothing parameter
  
  # Convert spline results to a data frame for ggplot
  spline_df <- data.frame(
    feature = spline_fit$x,
    yhat_smooth = spline_fit$y
  )
  colnames(spline_df)[1] <- feature
  # Create a proper ggplot
  print(ggplot(spline_df, aes_string(x = feature, y = "yhat_smooth")) +
    geom_line() +
    ggtitle(paste("Partial Dependence of", feature)) +
    xlab(feature) +
    ylab("Predicted Maternal Mortality Rate") +
    theme_minimal())
}
```

- From the graphs we can see that :-

1. Maternal mortality rates decline as income increase and at lower income levels maternal mortality rate increase.

2. High poverty rates are strongly associated with increase maternal mortality 

3. Maternal Mortality is high in places with a large American Indian and African American population.

4. Increase distance to obsetric care results in higher mortality rate and lower mortaity rates if the distance is short 

5. A high availability of health care centers is associated with low maternal deaths.


#Overall Conclusion

## 1.Socioeconomic Factors :- 

- High household income and low poverty rate are associated with a reduced maternal mortality. This suggests that better economic conditions can help in controlling Maternal Mortality Rates.

## 2.Healthcare Factors :-

- Proximity to obsetric care and availability of specialized hospitals can help control maternal mortality rates.

## 3. Demographic Factors :-

- Black and Indian population face more risks showing the need to fix unfair healthcare access and quality.


#REFERENCES

- Agency for Healthcare Research and Quality (AHRQ) Social Determinants of Health Database (2019-2020).

- R packages : ranger, ggplot2, randomforest etc

- R tools used : predict, lm etc
 






