---
title: "Project-6101"
author: "Sunny"
date: "2024-10-09"
output: 
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
---
```{r init, include=FALSE}
library(ezids)
library(dplyr)
library(ggplot2)
library(corrplot)
```




```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, results = "markup")
options(scipen=999, digits = 5) 
```

# Data
I just select the columns "Median household income (dollars, inflation-adjusted to data file year)
", "Total number of maternal deaths per 100,000 population" and "Total number of drug overdose deaths involving any opioid per 100,000 population".

```{r}
data_raw <- read.csv("C:/Users/qluo/Box/6101/project/GROUP-PROJECT-1/Combined_SDOH_2019_2020.csv")
data2 <- subset(data_raw, select = c("ACS_MEDIAN_HH_INC" ,"CDCW_MATERNAL_DTH_RATE", "CDCW_OPIOID_DTH_RATE"))
colnames(data2) <- c("Median_household_income",
                     "Total_number_of_maternal_deaths",
                     "Total_number_of_drug_overdose_deaths")
data2 <- na.omit(data2)
```


```{r}
summary(data2)
```

# EDA

First, the histograms.

```{r }
ggplot( data=data2, aes(Median_household_income))+
  geom_histogram(binwidth=5000, col="black", fill="red",boundary=35000)+
  labs(x="Median household income", 
       y= "Frequency", 
       title="Histogram of Median household income")
ggplot(data=data2,aes(Total_number_of_maternal_deaths))+
  geom_histogram(breaks=seq(0,2.6,by=.1),
                 col="black",
                 fill="green")+
  labs(x="Total number of maternal deaths( per 100,000 population)",
       y="Frequency",
       title="Histogram of Total number of maternal deaths")
ggplot(data=data2,aes(Total_number_of_drug_overdose_deaths))+
  geom_histogram(breaks=seq(0,110,by=5),
                 col="black",
                 fill="blue")+
  labs(x="Total number of drug overdose deaths( per 100,000 population)",
       y="Frequency",
       title="Histogram of Total number of drug overdose deaths")
```

Second, the Q-Q plot:

```{r}
qqnorm(data2$Median_household_income,main="Q-Q plot of Median household income")
qqline(data2$Median_household_income)
qqnorm(data2$Total_number_of_maternal_deaths,main="Q-Q plot of Total number of maternal deaths")
qqline(data2$Total_number_of_maternal_deaths)
qqnorm(data2$Total_number_of_drug_overdose_deaths,main="Q-Q plot of Total number of drug overdose deaths")
qqline(data2$Total_number_of_drug_overdose_deaths)

```



Now, boxplots.

```{r}
data2$Income_quartile <- ntile(data2$Median_household_income,4)
data2$Income_label <- case_when(
  data2$Income_quartile==1 ~ "Low Income",
  data2$Income_quartile==2 ~ "Middle Income",
  data2$Income_quartile==3 ~ "Upper Middle Income",
  data2$Income_quartile==4 ~ "High Income"
)
data2$Income_label <- factor(data2$Income_label,
                             level=c("Low Income", "Middle Income", "Upper Middle Income", "High Income"))
ggplot(data2,aes(x=Income_label,
                 y=Total_number_of_maternal_deaths))+
  geom_boxplot(col="black", fill="orange",outlier.color = "red")+
  labs(x="Median Household Income quartile", 
       y= "Frequency", 
       title="Boxplot of Maternal Deaths by Median household Income Quartile")

ggplot(data2, aes(x=Income_label,
                  y=Total_number_of_drug_overdose_deaths))+
  geom_boxplot(col="black",fill="blue",outlier.color = "red")+
  labs(x="Median Household Income quartile", 
       y= "Frequency", 
       title="Boxplot of Drug Overdose Deaths by Median household Income Quartile")
```

The total number of maternal deaths varies significantly between different income groups; however, the pattern in the number of drug overdose deaths is less distinct when viewed through the boxplots.

# ANOVA Test
Let's see if the Total number of maternal deaths per 100,000 population and Total number of drug overdose deaths per 100,000 population are different among different Income groups.

```{r}
anova_maternal_deaths <- aov(Total_number_of_maternal_deaths ~ Income_label, data=data2)
anova_drug_overdose_deaths <- aov(Total_number_of_drug_overdose_deaths ~ Income_label, data=data2)
xkabledply(anova_maternal_deaths)
xkabledply(anova_drug_overdose_deaths)

```


From the above ANOVA tests, both the total number of maternal deaths and drug overdose deaths are not all the same among different income groups. Let's do Post-hoc Tukey HSD to see which groups are different from others.


## Post-hoc Tukey HSD 

```{r }
tukey_maternal_deaths <- TukeyHSD(anova_maternal_deaths)
tukey_drug_overdose_deaths <- TukeyHSD(anova_drug_overdose_deaths)
tukey_maternal_deaths
tukey_drug_overdose_deaths
```

From the Post-hoc Tukey HSD, the total number of maternal deaths of all the groups are different from others except the Upper Middle Income and Middle Income at 5% significance level. However, only the High Income and Middle Income are different in the total number of drug overdose deaths at the 5% significance level. 

Let's have a look at the correlation coefficients between these variables.

# Correlation Matrix

```{r}
#install.packages("corrplot")
library("corrplot")
cor_matrix <- cor(data2[,c("Median_household_income","Total_number_of_maternal_deaths",
                           "Total_number_of_drug_overdose_deaths")])
cor_matrix
```


The correlation coefficient between income and maternal death is -0.5052, suggesting the negative relationship between the two variables.